\bf The forestecology R package for fitting and assessing models of interspecies competitive effects on the growth of trees

### 2021-02-22 paper/forestecology.tex
Done second pass of Section 2: Example

\if0\blind
{
  \title{\bf The forestecology R package for fitting and assessing models of
interspecies competition on the growth of trees}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf The forestecology R package for fitting and assessing models of
interspecies competition on the growth of trees}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
1. When modeling growth of trees forest ecologists often incorporate the
effect of interspecies competition. Many such models are based on a
neighborhood effect assumption whereby all trees within a fixed distance
of all focal trees are considered competitors. Methods are needed to
evaluate the effect of interspecies competition and to assess their
quality.\\
2. We present the \texttt{forestecology} package providing methods for
both 1) evaluating the out-of-sample performance of our model using
spatial-crossvaliation and 2) testing a null hypothesis of of no impact
of competitor species' identity on the growth of trees using a
permutation test. We implement a class and methods using R's S3
object-oriented system, for a specific linear, Bayesian neighborhood
competition model of tree growth.\\
3. We demonstrate the package's functions using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO network of reseach sites. Given ForestGEO's data
collection protocols and data formatting standards, the package
cross-compatibility of code. We show both that 1) competitor species
identity matters and 2) that not spatially cross-validating leads to
error estimates that are overly optimistic.\\
4. The package follows \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence to
intuitively display the sequence of steps of analysis from start to
finish. Additionally, most inputs/outputs of functions assume an are of
\texttt{sf} class from the simple features package, thereby facilitating
all wrangling and visualization of geospatial data. Lastly, even though
our package is currently limited to one specific model, the package is
setup such that it can be easily extended to other models.
\end{abstract}

@@

We demonstrate the \texttt{forestecology} package's features on the
Smithsonian Conservation Biology Institute (SCBI) large forest dynamics
plot, located at the Smithsonian's National Zoo and Conservation Biology
Institute in Front Royal, VA, USA. The 25.6 ha (640 x 400 m) plot is
located at the intersection of three of the major physiographic
provinces of the eastern US: the Blue Ridge, Ridge and Valley, and
Piedmont provinces and is adjacent to the northern end of Shenandoah
National Park. The forest type is typical mature secondary eastern mixed
deciduous forest, with a canopy dominated by tulip poplar
(\emph{Liriodendron tulipifera}), oaks (\emph{Quercus} spp.), and
hickories (\emph{Carya} spp.), and an understory composed mainly of
spicebush (\emph{Lindera benzoin}), paw-paw (\emph{Asimina triloba}),
American hornbeam (\emph{Carpinus caroliniana}), and witch hazel
(\emph{Hamamelis virginiana}) \citet{bourg_initial_2013}.

There are four steps in our modeling pipeline:

\begin{itemize}
\tightlist
\item
  Step 1: Compute the growth of trees based on census data.
\item
  Step 2: Add spatial information:

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    Define buffer region trees.
  \item
    Add spatial cross-validation block information.
  \end{enumerate}
\item
  Step 3: Identify all focal and corresponding competitor trees.
\item
  Step 4: Fit model, as well as:

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    Compute fitted/predicted values.
  \item
    Visualize posterior distributions.
  \end{enumerate}
\end{itemize}

Using these four steps, we can address the two following questions of
ecological interest:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Evaluate the effect of competitor species identity using permutation
  tests.
\item
  Evaluate model performance using spatial cross-validation.
\end{enumerate}

We load all necessary packages.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\KeywordTok{library}\NormalTok{(lubridate)}
\KeywordTok{library}\NormalTok{(sf)}
\KeywordTok{library}\NormalTok{(forestecology)}
\KeywordTok{library}\NormalTok{(blockCV)}
\KeywordTok{library}\NormalTok{(patchwork)}
\end{Highlighting}
\end{Shaded}

@@

The first step in the our analysis sequence is to compute the growth of
trees using data from two censuses. The \texttt{compute\_growth()}
function computes growth assuming census data that follows ForestGEO
standards. Despite such standards, minor variations will still exist
between sites thereby necessitating some data wrangling and checking.
For example, the SCBI site records all DBH's in millimeters, whereas the
Michigan Big Woods site records them in centimeters
\citet{andersonteixeira_ctfs-forestgeo_2015}
\citet{allen_michigan_2020}. The data format of other sites may be such
that our \texttt{compute\_growth()} function doesn't work at all.
However, in the end all that matters is that the growth of all trees is
saved in a data frame of class \texttt{sf} whereby the geolocation of
each tree is presented in a \texttt{geometry} variable of type
\texttt{\textless{}POINT\textgreater{}} and at a minimum the data
contains the following variables: a variable uniquely identifying each
tree-stem, \texttt{sp} of type \texttt{fct} factor identifying species,
\texttt{dbh1} and \texttt{dbh2} of type \texttt{dbl} quantifying the DBH
at earlier and later census, and \texttt{growth} of type \texttt{dbl}
double quantifying the average annual growth in centimeters.

We load both 2008 and 2014 SCBI census data \texttt{.csv} files as they
existed on GitHub on November 20, 2020. After selecting only relevant
variables, we perform a few additional data wrangling steps: convert the
variable with the date of measurement to be of type \texttt{date},
convert DBH to be in centimeters\footnote{A rule of thumb to determine
  the units of DBH is check if the smallest non-zero and non-missing
  measurement is 1 or 10. If the former, then centimeters. If the later,
  then millimeters. This is because ForestGEO protocols state that only
  trees with DBH greater or equal to 1cm should be included in censuses.},
convert the \texttt{sp} variable containing species information from
type \texttt{chr} character to \texttt{fct} factor (we will discuss the
need for this in Section \ref{spatial-cross-validation}). Furthermore,
in order to speed up computation for purposes of this example, we only
consider a 9 ha subsection of the 25.6 ha of the SCBI site: \texttt{gx}
from 0--300 instead of 0--400 and \texttt{gy} from 300--600 instead of
0--640.

@@

These two data frames are then used as the two primary arguments to the
\texttt{compute\_growth()} function, along with the \texttt{id} argument
whereby the user specifies the name of the variable that uniquely
identifies each tree-stem under consideration (note this does not
include resprouts in the later census):

@@

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes a numerical \texttt{growth} reflecting the
average annual growth in DBH (in cm) for all trees that were alive at
both time points as well a \texttt{geometry} variable encoding each
tree's geolocation. Furthermore, variables that (in theory) remain
unchanged between censuses appear only once, such as location variables
\texttt{gx} and \texttt{gy}; as well as species-related variables.
Variables that should change between censuses are suffixed with
\texttt{1} and \texttt{2} indicating the earlier and later censuses,
such as \texttt{dbh1/dbh2} and \texttt{codes1/codes2}.

@@

TODO: Rescale points in this plot:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ growth_scbi, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =}\NormalTok{ growth)) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_size}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.01}\NormalTok{, }\FloatTok{0.1}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{range =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.1}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\caption{Step 1 - Compute growth of trees based on census data: All trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot.}\label{fig:scbi-trees}
\end{figure}

@@

The next step in our analysis sequence is to add spatial information to
our main \texttt{growth\_scbi} data frame. The first element of spatial
information we add is a ``buffer region'' to the periphery of the study
region. Since some of our model's explanatory variables such as
competitor basal area are cumulative, we must ensure that all trees
being modeled are not biased to have different neighbor structures. This
is of particular concern for trees at the boundary of study regions,
which will not have the same number of neighbors that act as competitors
as trees in the internal part of the study region. In order to account
for such edge effects only trees who are not part of this buffer region,
i.e.~are part of the interior of the study region, will have their
growths modeled \citet{waller_applied_2004}.

Our model of interspecific competition relies on a spatial definition of
who the competitor trees are for focal trees of interest: all trees
within a distance \texttt{comp\_dist} of a focal tree are considered its
competitors (assuming the same units as the \texttt{gx} and \texttt{gy}
location variables). In our case we set this value below at 7.5m, a
value informed by \citet{canham_neighborhood_2004}
\citet{uriarte_spatially_2004} \citet{canham_neighborhood_2006}. Using
this value along with a manually constructed \texttt{sf} object
representation of the study region's boundary, we apply the
\texttt{add\_buffer\_variable()} to our \texttt{growth\_scbi} data frame
to add a \texttt{buffer} boolean variable: all trees who have
\texttt{buffer} set to \texttt{FALSE} will be our focal trees whose
growths are modeled, whereas those with \texttt{buffer} set to
\texttt{TRUE} will only be considered as competitor trees whose growth
will not be modeled.

@@

The second element of spatial information are blocks corresponding to
folds of a spatial cross-validation algorithm used to estimate model
error. Conventional cross-validation algorithms assign observations to
folds by randomly resampling individual observations. However,
underlying this algorithm is an assumption that the observations are
independent. In the case of forest census data, observations exhibit
spatial autocorrelation. This spatial dependence is incorporated into
the cross-validation algorithm by randomly resampling spatial blocks of
trees \citet{roberts_cross-validation_2017}
\citet{pohjankukka_estimating_2017}. We therefore associate each
observed tree to one of \(k\) spatial folds. In the example below, we
first manually define two folds that partition the study region as an
\texttt{sf} object. We then use the output of the
\texttt{spatialBlock()} function from the \texttt{blockCV} package to
associate each tree in \texttt{growth\_scbi} to the correct fold (saved
in the \texttt{foldID} variable) \citet{valavi_blockcv_2019}.
\textbackslash footnote\{In the Appendix we present an example where the
folds themselves are also created using the \texttt{spatialBlock()}
function given a specified \texttt{cv\_block\_size}.\}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Manually define spatial blocks to act as folds}
\NormalTok{fold1 <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{300}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{300}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{450}\NormalTok{))}
\NormalTok{fold2 <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{300}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{300}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{450}\NormalTok{))}
\NormalTok{fold3 <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{600}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{600}\NormalTok{))}
\NormalTok{fold4 <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{600}\NormalTok{))}
\NormalTok{n_fold <-}\StringTok{ }\DecValTok{4}

@@

\caption{Step 2 - Add spatial information: Buffer region and spatial cross-validation blocks (1 through 4). All trees in the interior of the study region will be the focal trees whose growth will modeled.}\label{fig:scbi-spatial-information}
\end{figure}

@@

The next step in our analysis sequence is to identify all focal trees
and their corresponding competitor trees. The
\texttt{create\_focal\_vs\_comp()} functions performs these tasks and
returns a new data frame of type \texttt{sf} containing this
information. On top of the previously discussed arguments
\texttt{comp\_dist} defining the competition neighborhood and
\texttt{id} indicating which variable in the data frame uniquely
identifies each tree-stem, this function also requires an \texttt{sf}
object representation of the spatial cross-validation blocks/folds; in
our case, this was manually encoded in the \texttt{blocks\_scbi} in
Section \ref{spatial-information} while in our Appendix we present an
example where this was performed using \texttt{spatialBlock()} from the
\texttt{blockCV} package. We present the resulting data frame below with
the \texttt{foldID} variable omitted for compactness of presentation.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{growth_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{create_focal_vs_comp}\NormalTok{(comp_dist, }\DataTypeTok{cv_grid_sf =}\NormalTok{ blocks_scbi, }\DataTypeTok{id =} \StringTok{"stemID"}\NormalTok{)}
\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{foldID)}
\CommentTok{## # A tibble: 6,296 x 6}
\CommentTok{##    focal_ID focal_sp   dbh     geometry  growth comp             }
\CommentTok{##       <dbl> <fct>    <dbl>      <POINT>   <dbl> <list>           }
\CommentTok{##  1        4 nysy     13.6  (14.2 428.5)  0.103  <tibble [20 x 4]>}
\CommentTok{##  2        5 havi      8.8   (9.4 436.4)  0.150  <tibble [32 x 4]>}
\CommentTok{##  3       79 tiam     47.7    (40 381.1) -0.161  <tibble [20 x 4]>}
\CommentTok{##  4       80 caca      5.15 (38.7 421.7)  0.253  <tibble [12 x 4]>}
\CommentTok{##  5       96 libe      2.3      (60 310)  0.262  <tibble [14 x 4]>}
\CommentTok{##  6      101 litu     65.4  (47.1 567.3)  0.552  <tibble [19 x 4]>}
\CommentTok{##  7      102 astr      1.99 (40.8 575.5)  0.0954 <tibble [44 x 4]>}
\CommentTok{##  8      126 cato     37.4  (60.6 400.2)  0.165  <tibble [16 x 4]>}
\CommentTok{##  9      127 caca      8.72 (72.7 514.1)  0.0370 <tibble [14 x 4]>}
\CommentTok{## 10      139 astr      1.71 (96.7 315.1)  0.0549 <tibble [48 x 4]>}
\CommentTok{## # ... with 6,286 more rows}
\end{Highlighting}
\end{Shaded}

TODO: Below reconcile the number of rows as they off by one from
\texttt{growth\_scbi} pipe \texttt{filter(!is.na(growth)\ \&\ !buffer)}.
perhaps by removing NA's in the \texttt{growth\_scbi} stage.

The resulting data frame \texttt{focal\_vs\_comp\_scbi} has 6296 rows,
representing the subset of the 7954 trees in \texttt{growth\_scbi} that
will be considered as focal trees and thus have their growths modeled.
Recall from Section \ref{spatial-information} this consists all trees
that are not part of the buffer region in Figure
\ref{fig:scbi-spatial-information}. Two new variables \texttt{focal\_ID}
and \texttt{focal\_sp} related to tree-stem identification and species
information. Most notably however is a new variable \texttt{comp} which
contains information on all competitor trees for a given focal tree
saved in list-column format, a feature of the \texttt{tidyr} package
\citet{tidyr_package}. For example, we drill-down on the tree with
\texttt{focal\_ID} equal to 4. It has 20 competitor trees each described
by 4 variables as indicated by the fact that \texttt{comp} is a
\texttt{\textless{}tibble\ {[}20\ ×\ 4{]}\textgreater{}}.

@@

\caption{Step 3 - Identify all focal and corresponding competitor trees: All 20 competitor trees of focal tree 4.}\label{fig:scbi-focal-vs-comp-map}
\end{figure}

>>>>

\if0\blind
{
  \title{\bf The forestecology R package for fitting and assessing models of
interspecies competitive effects on the growth of trees}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf The forestecology R package for fitting and assessing models of
interspecies competitive effects on the growth of trees}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
1. Many models for the growth of trees that incorporate the effect of
interspecies competition are based on a neighborhood effect assumption
whereby all trees within a fixed distance of all focal trees are
considered competitors. Methods and tools are needed to quantify this
competitive effect and assess the quality of all resulting models\\
2. We present the \texttt{forestecology} package providing methods for
both 1) evaluating the effect of competitor species identity using
permutation tests and 2) evaluating model performance using spatial
cross-validation. Following \citet{allen_permutation_2020}, we implement
a Bayesian linear regression competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We demonstrate that both
1) competitor species identity matters and 2) that not spatially
cross-validating leads to error estimates that are overly optimistic.\\
4. The package features 1) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, 2)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and 3) R S3 object-oriented implementation of
the Bayesian linear regression model. These three facts allow for clear
articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

@@

We demonstrate the \texttt{forestecology} package's functionality on
data from the Smithsonian Conservation Biology Institute (SCBI) large
forest dynamics plot, located at the Smithsonian's National Zoo and
Conservation Biology Institute in Front Royal, VA, USA. The 25.6 ha (640
x 400 m) plot is located at the intersection of three of the major
physiographic provinces of the eastern US: the Blue Ridge, Ridge and
Valley, and Piedmont provinces and is adjacent to the northern end of
Shenandoah National Park. The forest type is typical mature secondary
eastern mixed deciduous forest, with a canopy dominated by tulip poplar
(\emph{Liriodendron tulipifera}), oaks (\emph{Quercus} spp.), and
hickories (\emph{Carya} spp.), and an understory composed mainly of
spicebush (\emph{Lindera benzoin}), paw-paw (\emph{Asimina triloba}),
American hornbeam (\emph{Carpinus caroliniana}), and witch hazel
(\emph{Hamamelis virginiana}) \citet{bourg_initial_2013}.

The \texttt{forestecology} package attempts has the following two
ecological goals: 1) to evaluate the effect of competitor species
identity using permutation tests and 2) to evaluate model performance
using spatial cross-validation. To achieve these two goals, we outline a
basic analysis sequence comprising of these four main steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute the growth of trees based on census data.
\item
  Add spatial information:

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    Define a buffer region of trees.
  \item
    Add spatial cross-validation block information.
  \end{enumerate}
\item
  Identify all focal trees and their competitors.
\item
  Apply model, which includes:

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    Fit model.
  \item
    Compute fitted/predicted values.
  \item
    Visualize posterior distributions.
  \end{enumerate}
\end{enumerate}

We start by loading all necessary packages.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}
\KeywordTok{library}\NormalTok{(lubridate)}
\KeywordTok{library}\NormalTok{(sf)}
\KeywordTok{library}\NormalTok{(patchwork)}
\KeywordTok{library}\NormalTok{(forestecology)}
\KeywordTok{library}\NormalTok{(blockCV)}
\end{Highlighting}
\end{Shaded}

@@

The first step in the our analysis sequence is to compute the growth of
trees using data from two censuses. The \texttt{compute\_growth()}
function computes growth assuming census data that roughly follows
ForestGEO standards. Despite such standards, minor variations will still
exist between sites, thereby necessitating some data wrangling and
checking. For example, the SCBI site records all DBH's in millimeters,
whereas the Michigan Big Woods site records them in centimeters
\citet{andersonteixeira_ctfs-forestgeo_2015}
\citet{allen_michigan_2020}.

We load both 2008 and 2014 SCBI census data \texttt{.csv} files as they
existed on GitHub on November 20, 2020. After selecting only the
relevant variables, we perform a few additional data wrangling steps:
convert the character variable with the date of measurement to be of
explicit type \texttt{date}, convert DBH to be in centimeters\footnote{A
  rule of thumb to determine the units of DBH is to check if the
  smallest non-zero and non-missing measurement is 1 or 10. If the
  former, then centimeters. If the later, then millimeters. This is
  because ForestGEO protocols state that only trees with DBH greater or
  equal to 1cm should be included in censuses.}, convert the \texttt{sp}
variable containing species information from type \texttt{chr} character
to \texttt{fct} factor.\footnote{In our spatial cross-validation
  algorithm in Section \ref{spatial-cross-validation} issues can occur
  when rare species do not occur in the training set, but then are
  encountered in the test set. This risk is mitigated by representing
  \texttt{sp} as a factor variable, which has a complete list of all
  levels of the categorical variable.}Furthermore, in order to speed up
computation for purposes of this example, we only consider a 9 ha
subsection of the 25.6 ha of the SCBI site: \texttt{gx} from 0--300
instead of 0--400 and \texttt{gy} from 300--600 instead of 0--640.

@@

These two data frames are then used as arguments to the
\texttt{compute\_growth()} function, along with the \texttt{id} argument
that specifies the variable that uniquely identifies each tree-stem.
Note furthermore that we discard all resprouts in the later census
(those with \texttt{code\ ==\ R}).

@@

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes \texttt{growth}, the average annual growth in
DBH (in cm) for all trees that were alive at both time points, as well a
\texttt{geometry} variable encoding each tree's geolocation.
Furthermore, the variables that should remain unchanged between censuses
appear only once, such as location variables \texttt{gx} and
\texttt{gy}; as well as species-related variables. Variables that should
change between censuses are suffixed with \texttt{1} and \texttt{2}
indicating the earlier and later censuses, such as \texttt{dbh1/dbh2}
and \texttt{codes1/codes2}.

The data format of other sites may be such that our
\texttt{compute\_growth()} function doesn't work at all. However, in the
end all that matters is that the growth of all trees is saved in a data
frame of class \texttt{sf} whereby the geolocation of each tree is
presented in a \texttt{geometry} variable of type
\texttt{\textless{}POINT\textgreater{}} and at a minimum the data
contains the variables above.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# }\AlertTok{TODO}\CommentTok{: Rescale points in this plot}
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ growth_scbi, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =}\NormalTok{ growth)) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_size}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.01}\NormalTok{, }\FloatTok{0.1}\NormalTok{, }\DecValTok{1}\NormalTok{), }\DataTypeTok{range =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.1}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\caption{Compute growth of trees based on census data: Map with growth of all trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot.}\label{fig:scbi-trees}
\end{figure}

@@

The next step in our analysis sequence is to add spatial information to
our main \texttt{growth\_scbi} data frame. The first element of spatial
information we add is a ``buffer region'' to the periphery of the study
region. Since some of our model's explanatory variables are cumulative
(such as competitor basal area), we must ensure that all trees being
modeled are not biased to have different neighbor structures. This is of
concern for trees at the boundary of study regions, which will not have
the same number of neighbors as trees in the internal part of the study
region. In order to account for such edge effects, only trees that are
not part of this buffer region, i.e.~are part of the interior of the
study region, will have their growths modeled
\citet{waller_applied_2004}.

Our model of interspecific competition relies on a spatial definition of
who the competitor trees are for focal trees of interest: all trees
within a distance \texttt{comp\_dist} of a focal tree are considered its
competitors (assuming the same units as the \texttt{gx} and \texttt{gy}
location variables). In our case we set this value at 7.5m, a value
informed by \citet{canham_neighborhood_2004}
\citet{uriarte_spatially_2004} \citet{canham_neighborhood_2006}. Using
this value along with a manually constructed \texttt{sf} object
representation of the study region's boundary, we apply the
\texttt{add\_buffer\_variable()} to our \texttt{growth\_scbi} data frame
to add a \texttt{buffer} boolean variable: all trees who have
\texttt{buffer} set to \texttt{FALSE} will be our focal trees whose
growths are modeled, whereas those with \texttt{TRUE} will only be
considered as competitor trees whose growth will not be modeled.

@@

The second element of spatial information are blocks corresponding to
folds of a spatial cross-validation algorithm used to estimate
out-of-sample model error. Conventional cross-validation algorithms
assign observations to folds by randomly resampling individual
observations. However, an assumption that the observations are
independent underlies any cross-validation algorithm. In the case of
forest census data, observations exhibit spatial autocorrelation. We
therefore incorporate this spatial dependence into the cross-validation
algorithm by randomly resampling spatial blocks of trees
\citet{roberts_cross-validation_2017}
\citet{pohjankukka_estimating_2017}.

In the example below, we first manually define four folds that partition
the study region as an \texttt{sf} object. We then use the output of the
\texttt{spatialBlock()} function from the \texttt{blockCV} package to
associate each tree in \texttt{growth\_scbi} to the correct fold (saved
in the \texttt{foldID} variable) \citet{valavi_blockcv_2019}. \footnote{In
  the Appendix we present an example where the folds themselves are also
  created using the \texttt{spatialBlock()} function given a specified
  \texttt{cv\_block\_size}.}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Manually define spatial blocks to act as folds}
\NormalTok{n_fold <-}\StringTok{ }\DecValTok{4}
\NormalTok{fold1 <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{300}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{300}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{450}\NormalTok{))}
\NormalTok{fold2 <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{300}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{300}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{450}\NormalTok{))}
\NormalTok{fold3 <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{600}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{600}\NormalTok{))}
\NormalTok{fold4 <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{450}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{), }\KeywordTok{c}\NormalTok{(}\DecValTok{150}\NormalTok{, }\DecValTok{600}\NormalTok{))}

@@

\caption{Add spatial information: Buffer region and spatial cross-validation blocks (1 through 4). All trees in the interior of the study region (i.e. not part of buffer) will be the focal trees whose growth will be modeled.}\label{fig:scbi-spatial-information}
\end{figure}

@@

The next step in our analysis sequence is to identify all focal trees
and their corresponding competitor trees. The
\texttt{create\_focal\_vs\_comp()} functions performs these tasks and
returns a new data frame of type \texttt{sf}. On top of the previous
arguments \texttt{comp\_dist} defining the competition neighborhood and
\texttt{id} indicating which variable uniquely identifies each
tree-stem, this function also requires an \texttt{sf} object
representation of the spatial cross-validation blocks/folds; in our
case, this was manually encoded in the \texttt{blocks\_scbi} in Section
\ref{spatial-information} while in our Appendix we present an example
where this was performed using \texttt{spatialBlock()} from the
\texttt{blockCV} package. We present the resulting data frame below with
the \texttt{foldID} variable omitted for compactness of presentation.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# }\AlertTok{TODO}\CommentTok{: Below reconcile the number of rows as they off by one from growth_scbi %>% }
\CommentTok{# filter(!is.na(growth) & !buffer). Perhaps by removing NA's in the growth_scbi stage?}
\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{growth_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{create_focal_vs_comp}\NormalTok{(comp_dist, }\DataTypeTok{cv_grid_sf =}\NormalTok{ blocks_scbi, }\DataTypeTok{id =} \StringTok{"stemID"}\NormalTok{)}
\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{foldID)}
\CommentTok{## # A tibble: 6,296 x 6}
\CommentTok{##    focal_ID focal_sp   dbh     geometry  growth comp             }
\CommentTok{##       <dbl> <fct>    <dbl>      <POINT>   <dbl> <list>           }
\CommentTok{##  1        4 nysy     13.6  (14.2 428.5)  0.103  <tibble [20 x 4]>}
\CommentTok{##  2        5 havi      8.8   (9.4 436.4)  0.150  <tibble [32 x 4]>}
\CommentTok{##  3       79 tiam     47.7    (40 381.1) -0.161  <tibble [20 x 4]>}
\CommentTok{##  4       80 caca      5.15 (38.7 421.7)  0.253  <tibble [12 x 4]>}
\CommentTok{##  5       96 libe      2.3      (60 310)  0.262  <tibble [14 x 4]>}
\CommentTok{##  6      101 litu     65.4  (47.1 567.3)  0.552  <tibble [19 x 4]>}
\CommentTok{##  7      102 astr      1.99 (40.8 575.5)  0.0954 <tibble [44 x 4]>}
\CommentTok{##  8      126 cato     37.4  (60.6 400.2)  0.165  <tibble [16 x 4]>}
\CommentTok{##  9      127 caca      8.72 (72.7 514.1)  0.0370 <tibble [14 x 4]>}
\CommentTok{## 10      139 astr      1.71 (96.7 315.1)  0.0549 <tibble [48 x 4]>}
\CommentTok{## # ... with 6,286 more rows}
\end{Highlighting}
\end{Shaded}

The resulting data frame \texttt{focal\_vs\_comp\_scbi} has 6296 rows,
representing the subset of the 7954 trees in \texttt{growth\_scbi} that
will be considered as focal trees. Two new variables \texttt{focal\_ID}
and \texttt{focal\_sp} relate to tree-stem identification and species
information. Most notably however is a new variable \texttt{comp} which
contains information on all competitor trees for a given focal tree,
saved in \texttt{tidyr} package list-column format
\citet{tidyr_package}. For example, we drill-down on the tree with
\texttt{focal\_ID} 4, which has 20 competitor trees each described by 4
variables as indicated by the fact that \texttt{comp} is a
\texttt{\textless{}tibble\ {[}20\ ×\ 4{]}\textgreater{}}.

@@

\caption{Identify all focal and corresponding competitor trees: All 20 competitor trees of focal tree 4.}\label{fig:scbi-focal-vs-comp-map}
\end{figure}
